Convergence Studies and Effective Alphas for Disks with Gravitational Instabilities. \\ I. The Constant Cooling Case

### 2010-11-04 Paper_1.tex
test commit
Edits to Paper. 1st pass.
	modified:   Paper_1.tex
	new file:   figures/qcompareres.eps
	modified:   general.bib

\newcounter{subequation}

\usepackage{color}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{amsmath, amsthm, amssymb}

@@

Using 3D hydrodynamic simulations of a cooling protostellar disk undergoing gravitational instabilities (GIs), we compute the effective \citet{shakura1973} $\alpha$ due to gravitational stresses and compare them to predictions from an analytic local theory for thin disks by \citet{gammie2001}.  Our goal is to determine how accurately a locally defined alpha can characterize mass and angular momentum transport by GIs in disks.  We examine disks with an imposed constant global cooling time \citep{mejia2005}.  Models were computed with grids of different spatial resolution to investigate how the computed alpha is affected by numerical resolution.

@@

Gravitational instabilities (GIs) play an important, and at times dominant, role in driving the structural and thermal
evolution of  protostellar disks \citep[for a review, see][]{durisen2005}.
%In particular, torques arising from GIs provide a very effective means of redistributing matter and 
%angular momentum within these disks 
%\citep{Lod04, Lod05, Vor05, Vor06, Bol09}.  
Thermal processes play the primary role in regulating the amplitude and outcome of these instabilities \citep{pickett1998,pickett2000,mejia2005,nelson1998,nelson2000a}. A link between a disk's thermodynamic state and its susceptibility to GIs is provided by the Toomre $Q$-parameter \citep{toomre1981};  $Q = c_s \kappa / \pi G \Sigma$, where $c_s$ is the sound speed, $\kappa$ is the epicyclic frequency ($\sim$ the rotation frequency 
$\Omega$ in a nearly Keplerian disk), and $\Sigma$ is the disk surface mass density. For $Q \al$  1.5 to 1.7,  small density perturbations in a disk grow exponentially on a time scale comparable to the rotation period. These perturbations manifest themselves as multi-arm spirals with a predominantly trailing pattern that transports angular momentum outward by gravitational torques \citep[e.g.]{larson1984,boss1984,durisen1986,papaloizou1991, laughlin1994,nelson1998,pickett1998}. The long range nature of these torques cause nonlinear
spiral structures to develop over a broad range of radii \citep{laughlin1998,nelson1998,nelson2000a,pickett1998, pickett2000,pickett2003}.

Numerous studies have shown that thermal physics can control GI amplitudes by a {\it balance} of heating and cooling \citep[for example]{tomley1991,tomley1994,pickett1998,pickett2000, pickett2003,gammie2001,boss2002,rice2003b,mejia2005,boley2006,boley2007, stamatellos2008,cossins2009}.
Heating by GIs tends to increase $c_s$, thus increasing $Q$. If disk heating persists, the instability will shut off once $Q$ becomes sufficiently large.  However, if the disk cools radiatively, then heating and cooling can reach a balance at nonlinear wave amplitude. GI-activity can sustain this balance  at a relatively constant,
but unstable, value of $Q$ \citep{paczynski1978, lin1981,goldreich1965}.  
For sufficiently low cooling rates, the disk can exist in a state of quasi-equilibrium, where cooling is 
balanced by heating arising from the inward flow of matter and shocks induced by the GIs
\citep{gammie2001, lodato2004, rice2005, cossins2009, vorobyov2010}.

Fully nonlinear hydrodynamic simulations of low-$Q$ disks show that multiple
modes can become unstable in the linear regime \citep[e.g.]{nelson1998, pickett1998} {\bf additional refs?}.  Even if the initial growth is dominated by one mode, numerous modes usually appear in the nonlinear regime. As a consequence, if a disk achieves a quasi-steady balance between heating and cooling, then there tends to be power at all $m$-values resolved by a given numerical method \citep{mejia2005,boley2006}.  
In this  {\it asymptotic state}, GI-active disks develop overlapping density structures of different strengths, geometries and coherence,producing strongly nonlinear turbulence that pervades the entire disk \citep[e.g.]{pickett2003, mejia2005,boley2006}.
\citet{gammie2001} refers to this state as {\it gravitoturbulence}. Gravitoturbulence provides a possible bridge between the detailed physics of GIs and the viscous transport parameter $\alpha$.

\citet{shakura1973} first parameterized turbulent transport in a steady-state disk with the  parameter $\alpha$, where the source of the turbulence is unspecified. Although their original work was directed towards X-ray bright
accretion disks around black holes, this heuristic approach has proved fruitful in many astrophysical situations by allowing for analytic solutions and relatively easy numerical modeling over time-spans inaccessible to detailed modeling. In the $\alpha$ disk formalism, when the accretion rate $\dot{M}$ is constant in both time and radius, $\dot{M} = 3\pi\alpha c_s^2 \Sigma / \Omega$ where $c_s$ is the sound speed, $\Sigma$ the surface density, and $\Omega$ the rotational frequency. The $\alpha$ formalism makes no assumption about the nature of the angular momentum transport, only that {\it heating and cooling are dominated by local processes}; the validity of an $\alpha$-disk picture requires self-consistency of local energy dissipation.

Dissipative locality in GI-active disks has not been rigorously tested and remains an open question. 
Several studies support the idea that GI transport in real disks is, in many important respects, an intrinsically global phenomenon and cannot be properly treated by a local $\alpha$-like prescription \citep{laughlin1996, balbus1999, lodato2005, mejia2005, boley2006, cai2008}.  This almost certainly applies to the embedded phase of both high and low-mass stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate \citep[e.g.]{yorke1993, laughlin1994, yorke1999,vorobyov2005,vorobyov2006}, and disks observed in the embedded phase can be quite massive relative to their central stars \citep[e.g.]{osorio2003}. 
However, despite the long-distance nature of gravitational torques, theoretical and numerical results suggest that in protostellar disks where heating and cooling are in balance it may be valid to represent GI-induced transport by an {\it effective viscosity}, a local process \citep[e.g.]{gammie2001, rice2005, lodato2004}. Indeed, \citet{gammie2001} derived an effective alpha for the case of a thin 2D gravitationally unstable disk where the heating due to local turbulent dissipation caused by GIs was balanced by the local cooling rate.  Modifications to this derivation have since been proposed by {\bf (Tom: include other derivations and refs)}.   Recent simulations \citep[e.g.]{cossins2009, vorobyov2010} suggest that, at least in the case of imposed cooling, $\alpha_{eff}$ roughly converges to the value predicted by \citeauthor{gammie2001}.

@@

Because this is a polytropic disk with an idealized cooling time, only the ratio of disk to stellar mass, given here by 0.153, is specified.  In what follows, quantities specified with real physical units are derived assuming a 1 $M_\odot$ central star and a 0.153  $M_\odot$ disk. In this case, the ORP is defined at $r = 33$ AU and one ORP $\approx$ 303 years.  Models have $t_{cool} = 2$ ORP.   {\bf (Scott: Please check these numbers)}
The disk has an initial surface density $\Sigma \sim r^{ -1}$, and initial inner and outer radii of 2.3 and 40 AU, respectively.

Initial conditions for the disk's structure and thermodynamic state were set using an equilibrium star plus disk model generated using a modified \citet{hachisu1986} self-consistent field (SCF) relaxation method \citep{pickett1996,pickett2003,mejiaphd2004,mejia2005,cai2006}. Here  the density and angular momentum distributions are iteratively solved, using the specified equation of state,  until convergence is achieved. This  equilibrium disk was then seeded with small random perturbations to allow GIs to grow.  The resultant marginally unstable disk served as the $t = 0$ load for the hydro code.

The initially unstable disk model passed through several phases of evolution \citep[see also]{pickett2003, mejia2005}.  The initial load is unstable to the growth of axisymmetric structures. During this axisymmetric phase, 
which lasts several ORPs, the disk shrinks slightly, in an axisymmetric manner, until it becomes unstable to nonaxisymmetric perturbations.    When this condition is met, the disk undergoes a strong burst of GIs that predominantly manifest themselves in one or two discrete global spiral modes.  The disk expands violently during the burst phase, producing a significant rearrangement of the disk's mass distribution on a timescale of one or
two ORPs. This is then followed by a period of several ORPs where heating temporarily washes out some of the nonaxisymmetry in the disk.  Finally the disk settles into a quasi-steady, long-lived asymptotic state of sustained GI-activity over a large part of the disk, with an overall balance of heating and cooling.

Simulations were run with four different angular resolutions, with grid sizes in $(r,\phi,z)$ of  $(j_{max}, k_{max}, l_{max}) =$ (512,64,64), (512,128,64), (512,256,64), and (512,512,64). Because of the large computational overhead associated with starting each simulation from$t = 0$, the $l_{max} =$ 128 simulation was run through the axisymmetric, burst and transition phases. Simulations using $l_{max} =$ 64, 256, and 512 all begin by interpolating the $l_{max} =$ 128 simulation to higher or lower azimuthal resolution at 9.6 ORPs, near the end of the transition
phase. Although this strategy saves a large amount of computational time, it limits the scope of the analysis to the asymptotic phase and it introduces a perturbation to the disk which takes several ORPs to subside. By 12 ORPs, the perturbation has dissipated and all four disks have transitioned to the asymptotic state.  All simulations were 
run through $\sim 18$ ORPs.

Figure \ref{Final_Q} shows $Q$, as a function of radius, at $t \sim$ 18 ORP, while table \ref{tbl:ams} lists $Q_{avg}$, the value of $Q$ of for the asymptotic disk, time-averaged of 12 -- 19 ORP and spatially averaged over 10 -- 40 AU, as a function of radius for each resolution, 
{\bf (Scott: Figure 1 needs to be redone - I have extracted the 
plot from your chapter 3 and it only shows the 128 model.  We should show all four resolutions
or ditch this figure)}.

@@

\begin{equation}
A_m = \frac{(a_m^2 + b_m^2)^{1/2}}{\pi\int\rho_o \varpi {\rm d}\varpi {\rm d}z},
\end{equation}
where
\renewcommand{\theequation}{\arabic{equation}\alph{subequation}}
\setcounter{subequation}{2}
\begin{equation}
a_m = \int \rho \cos(m\phi)\varpi {\rm d}\varpi {\rm d}z {\rm d}\phi,
\end{equation}
\addtocounter{equation}{-1}
\addtocounter{subequation}{1}
\begin{equation}
b_m =\int \rho \sin(m\phi)\varpi {\rm d}\varpi {\rm d}z {\rm d}\phi.
\end{equation}
\renewcommand{\theequation}{\arabic{equation}}

Here  $\rho_o$ is the axisymmetric (zero-order)  component of the density. In what follows, $A_m$ is averaged over time to suppress fluctuations on the dynamic time scale due to the chaotic nature of GIs in the asymptotic  phase.
In particular, $\langle A_{tot} \rangle$ will represent the power over the radial range 10 -- 40 AU, integrated from $ m = 2$ to $l_{max}/2$ and averaged over the time range 12 -- 19 ORPs. The density distribution cannot be resolved in azimuth on angular scales less than $2\pi/l_{max}$, therefore $A_m$ is limited to modes $m \le l_{max}/2$.  
Because the central star is artificially fixed to the grid center, the $m = 1$ mode cannot be be realistically treated; it is excluded from $\langle A_{tot} \rangle$.

@@

Previous studies have shown that disks in the asymptotic state have power at all resolvable $m$-values  \citep[e.g.]{mejia2005,boley2006}.  Table \ref{tbl:ams} lists $\langle A_{tot} \rangle$ along with $\langle A_{2-7} \rangle  /  \langle A_{tot} \rangle$,  the fraction of the total power falling into low-order ($m = 2$ -- 7) modes, for each resolution. While the total power is approximately independent of resolution, $\langle A_{tot} \rangle = 1.90 \pm 0.07$, 
the modal distribution of this power shows marked differences. In particular, the fractional power in lower order modes decreases monotonically with increasing resolution.  Over three quarters of $\langle A_{tot} \rangle$ resides
in modes $m =$ 2 -- 7 for the $l_{max} = 64$ simulation, while approximately half resides in these modes when $l_{max} =$ 512.

The shift of power from lower to higher order modes as the resolution of the computational grid is increased arises from the fact that, while $\langle A_{tot} \rangle$ is approximately independent of $l_{max}$, the number of degrees of freedom available for this power, $l_{max}/2$, increases linearly with the resolution. Thus power is naturally spread from lower order to higher order modes as the available $m$-values increases. Indeed, even if the disk is dominated by one or two low-order modes at the time of outburst, this power cascades to higher-order modes \citep[for example]{laughlin1997, laughlin1998,laughlin1996}.

Given the resolution dependence of how power is distributed,  we must examine what resolution is sufficient to properly capture the evolution of these disks. Table \ref{tbl:ams} shows that while the fractional power in lower order modes decreases with each doubling of $l_{max}$, the amount of change in going from $l_{max} = 256$ to 512 
is significantly less than the change going from $l_{max} = 64$ to 128, with a strong suggestion that this diminution will not proceed much further with higher azimuthal resolution. Indeed, a  Richardson extrapolation {\bf (add reference)} of the tabulated values of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ suggests a limiting value of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle \sim$ 0.43 for increasingly higher resolutions. This is displayed graphically in Figure 4, where the fractional power in lower order modes is plotted against the  normalized grid 
spacing for each value of $l_{max}$. The grid spacing  is normalized such that spacing of the $l_{max}$ = 512 grid
equals unity.  Convergence occurs as the normalized spacing goes to zero. The Richardson extrapolation value  of $\langle A_{2-7}\rangle  / \langle A_{tot}\rangle  = 0.43$ is plotted at a normalized grid spacing  equal to zero. More than half of the modal power in a converged disk resides in higher-order modes, modes that help define a state of gravitoturbulence. The fact that the $l_{max} =$ 512 simulation has $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ near the convergence value lends some confidence in the use of the $l_{max} =$ 512 model in examining the convergence of $\alpha$.

@@

Coherent density structures in a disk produce gravitational torques. Low-order modes in $\Delta\rho / \rho$  have longer wavelengths and, if coherent in radius, have longer lever arms to produce torques. On the other hand, high-order modes have relatively short wavelengths which, particularly if they lack coherence, produce more localized effects. The high-order modes tend to produce pockets of local GI activity, which can cancel each other
over the whole disk, leading to the expectation that  gravitational torques in the disk will decrease as $l_{max}$ is increased.    This can be tested by examining the torque contributions from  various Fourier modes.

The torque {\bf C} on a cylindrical surface of the disk at radius $\varpi$ can be obtained by integrating the stress tensor $T$ over the surface of the cylinder \citep{lyndenbell1972}, i.e.,
\begin{equation}
{\rm {\bf C}} = \int {\bf r} \times T \cdot {\rm d}S.
\end{equation}
If the stress tensor includes only gravitational stresses, the surface integral can be replaced with the volume integral
\begin{equation}
{\rm \bf C} = \int \rho {\bf r} \times \nabla \Phi {\rm d}V,
\end{equation}
where $\Phi$ is the gravitational potential.  Here we are interested only in the $z$-component of torque,  
\begin{equation}
\label{torque1}
{\rm \bf C}_Z = \int \rho \frac{\partial\Phi} {\partial \phi} {\rm d}V,
\end{equation}
as only this component drives mass and angular momentum transfer. The torque contribution from each mode $m$ can be calculated by replacing $\rho$ in Equation (5) with the density distribution reconstructed from a single Fourier component
\begin{equation}
\rho_m = a_{\phi m} \cos(m\phi) + b_{\phi m} \sin(m\phi),
\end{equation}
where $a_{\phi m} =\slantfrac{1}{\pi} \int \rho \cos(m\phi){\rm d}\phi$ and $b_{\phi m} =\slantfrac{1}{\pi} \int \rho \sin(m\phi){\rm d}\phi$, and only the gravitational potential produced by the mass distribution given by $\rho_m$ is included in $\Phi$.
{\bf (Scott: Equation (6) is the form you use in your chapter 3, but it
does not appear to be correct; it does not yield a correct value for $m=0$)}

Figure 5 displays $\langle {\rm \bf C}_{Ztot} \rangle$, the time-averaged gravitational torque summed over all disk modes (excluding $m=1$), along with the torques summed over low-order modes, $\langle {\rm \bf C}_{Z(2-n)} \rangle$, where $n = $ 2, 3, 4, and 5, as a function of radius and azimuthal resolution. 
{\bf (Note to all: the version of Figure 5 included with this draft  shows the torques
time-averaged over for four different time-frames.  I did this because I wanted to see if the disk was 
"settling down" with time.  I think the figure suggests  that the 12-14 ORP disk is not
yet in an asymptotic state - or further from it than at later times - I'm not sure this is
a significant difference.  Looking forward to Figure 7, it can be argued that the
disk is reasonably settled, in a statistical sense, shortly after 12 ORP.
Dick has told me he only believes the 12-18 ORP result - Dick, I hope 
I'm not misquoting you).  
} 
The scale of the $y$-axis in each subplot is the same, with the exception of the $l_{max} = 64$ simulation. 
In this case the scale is doubled.  Two features are immediately apparent: 1) the total gravitational torque is smaller for higher resolution simulations, and 2) the total torque is dominated by low-order modes. Both reflect the fact that a while higher fraction of the total power in larger $l_{max}$ simulations is in higher order modes, these modes make only minimal contributions to the gravitational torque. As resolution increases, non-axisymmetric power is shifted from low-order global modes, which produce relatively large gravitational torques, to high-order local modes, which tend to cancel each other out when integrated over the whole disk.

>>>>

\usepackage{natbib}
\usepackage{color}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{grffile}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{amssymb}

@@

Using 3D hydrodynamic simulations of a cooling protostellar disk undergoing gravitational instabilities (GIs), we compute the effective \citet{shakura1973} $\alpha$ due to gravitational stresses and compare them to predictions from an analytic local theory for thin disks by \citet{gammie2001}.  Our goal is to determine how accurately a locally defined $\alpha$ can characterize mass and angular momentum transport by GIs in disks.  We examine disks with an imposed constant global cooling time \citep{mejia2005}.  Models were computed with grids of different spatial resolution to investigate how the computed $\alpha$ is affected by numerical resolution.

@@

Gravitational instabilities (GIs) play an important, {\bf SM: I'm not sure this is true in general. Perhaps reword to ``GIs can play an important role...and may dominate the structure for certain phases of evolution''}and at times dominant, role in driving the structural and thermal
evolution of  protostellar disks \citep[for a review, see][]{durisen2005}.
%In particular, torques arising from GIs provide a very effective means of redistributing matter and 
%angular momentum within these disks 
%\citep{Lod04, Lod05, Vor05, Vor06, Bol09}.  
Thermal processes play the primary role in regulating the amplitude and outcome of these instabilities \citep{pickett1998,pickett2000,mejia2005,nelson1998,nelson2000a}. A a disk's susceptibility to GIs is parameterized by the Toomre $Q$-parameter \citep{toomre1981};  $Q = c_s \kappa / \pi G \Sigma$, where $c_s$ is the sound speed, $\kappa$ is the epicyclic frequency ($\sim$ the rotation frequency 
$\Omega$ in a nearly Keplerian disk), and $\Sigma$ is the disk surface mass density. For $Q \al$  1.5 to 1.7,  small density perturbations in a disk grow exponentially on a time scale comparable to the rotation period. These perturbations manifest themselves as multi-arm spirals with a predominantly trailing pattern that transports angular momentum outward by gravitational torques \citep[e.g.]{larson1984,boss1984,durisen1986,papaloizou1991, laughlin1994,nelson1998,pickett1998}. The long range nature of these torques cause nonlinear
spiral structures to develop over a broad range of radii \citep{laughlin1998,nelson1998,nelson2000a,pickett1998, pickett2000,pickett2003}. {\bf SM: not sure of the meaning here. A low Q is what causes GIs to develop over a broad range of radii. Perhaps you mean the long range nature can cause coherent spiral structures to form with large radial extent?}

Numerous studies have shown that thermal physics control GI amplitudes by a {\it balance} of heating and cooling \citep[for example][]{tomley1991,tomley1994,pickett1998,pickett2000, pickett2003,gammie2001,boss2002,rice2003b,mejia2005,boley2006,boley2007, stamatellos2008,cossins2009}.
Heating by GIs tends to increase $c_s$, thus increasing $Q$. If disk heating persists, the instability will be suppressed once $Q$ becomes sufficiently large.  However, heating and cooling can reach a balance at nonlinear wave amplitude. GI-activity can sustain this balance  at a relatively constant,
but unstable, value of $Q$ \citep{paczynski1978, lin1981,goldreich1965},and 
the disk can exist in a state of quasi-equilibrium. In this state cooling is 
balanced by heating arising from the inward flow of matter and shocks induced by the GIs
\citep{gammie2001, lodato2004, rice2005, cossins2009, vorobyov2010}.

Fully nonlinear hydrodynamic simulations of low-$Q$ disks show that multiple
modes can become unstable in the linear regime \citep[e.g.][]{nelson1998, pickett1998} {\bf additional refs?}.  Even if the initial growth is dominated by one mode, numerous modes usually appear in the nonlinear regime. As a consequence, if a disk achieves a quasi-steady balance between heating and cooling, then there tends to be power at all $m$-values resolved by a given numerical method \citep{mejia2005,boley2006}.  
In this  {\it asymptotic state}, GI-active disks develop overlapping density structures of different strengths, geometries and coherence, producing strongly nonlinear turbulence that pervades the entire disk \citep[e.g.][]{pickett2003, mejia2005,boley2006}.
\citet{gammie2001} refers to this state as {\it gravitoturbulence}. Gravitoturbulence provides a possible bridge between the detailed physics of GIs and the viscous transport parameter $\alpha$.

\citet{shakura1973} first parameterized turbulent transport in a steady-state disk with the  parameter $\alpha$, where the source of the turbulence is unspecified. Although their original work was directed towards X-ray bright
accretion disks around black holes, this heuristic approach has proved fruitful in many astrophysical situations by allowing for analytic solutions and relatively easy numerical modeling over time-spans inaccessible to detailed modeling. In the $\alpha$ disk formalism, when the accretion rate $\dot{M}$ is constant in both time and radius, $\dot{M} = 3\pi\alpha c_s \Sigma H$ where $c_s$ is the sound speed, $\Sigma$ the surface density, and $H$ the disk scale height. The $\alpha$ formalism makes no assumption about the nature of the angular momentum transport, only that {\it heating and cooling are dominated by local processes}{\bf SM: This is not my understanding of the $\alpha$ formalism, I don't think it says anything about cooling. It also means that ang mom transport happens locally and that this causes local heating}; the validity of an $\alpha$-disk picture requires self-consistency of local energy dissipation.

Dissipative locality {\bf SM: need to explain what is meant by this more precisely} in GI-active disks has not been rigorously tested and remains an open question. 
Several studies support the idea that GI transport in real disks is, in many important respects, an intrinsically global phenomenon and cannot be properly treated by a local $\alpha$-like prescription \citep{laughlin1996, balbus1999, lodato2005, mejia2005, boley2006, cai2008}.  This almost certainly applies to the embedded phase of both high and low-mass stars when infall from the collapsing protostellar cloud feeds mass into the disk at a rapid
rate \citep[e.g.]{yorke1993, laughlin1994, yorke1999,vorobyov2005,vorobyov2006}, and disks observed in the embedded phase can be quite massive relative to their central stars \citep[e.g.][]{osorio2003}. 
However, despite the long-distance nature of gravitational torques, theoretical and numerical results suggest that in protostellar disks where heating and cooling are in balance it may be valid to represent GI-induced transport by an {\it effective viscosity}, a local process \citep[e.g.][]{gammie2001, rice2005, lodato2004}. Indeed, \citet{gammie2001} derived an effective alpha for the case of a thin 2D gravitationally unstable disk where the heating due to local turbulent dissipation caused by GIs was balanced by the local cooling rate.  Modifications to this derivation have since been proposed by {\bf (Tom: include other derivations and refs)}.   Recent simulations \citep[e.g.]{cossins2009, vorobyov2010} suggest that, at least in the case of imposed cooling, $\alpha_{\rm{eff}}$ roughly converges to the value predicted by \citeauthor{gammie2001}.

@@

Because this is a polytropic disk with an idealized cooling time, only the ratio of disk to stellar mass, given here by 0.153, is specified.  In what follows, quantities specified with real physical units are derived assuming a 1 $M_\odot$ central star and a 0.153  $M_\odot$ disk. In this case, the ORP is defined at $r = 33$ AU and one ORP $\approx$ 303 years.  Models have $t_{cool} = 2$ ORP. The disk has an initial surface density $\Sigma \sim r^{ -1}$, and initial inner and outer radii of 2.3 and 40 AU, respectively.

Initial conditions for the disk's structure and thermodynamic state were set using an equilibrium star plus disk model generated using a modified \citet{hachisu1986} self-consistent field (SCF) relaxation method \citep{pickett1996,pickett2003,mejiaphd2004,mejia2005,cai2006}. Here  the density and angular momentum distributions are iteratively solved, using the specified equation of state,  until convergence is achieved. This  equilibrium disk was then seeded with small 0.01\% $\delta\rho/\rho$ random perturbations to allow GIs to grow.  The resultant marginally unstable disk served as the initial disk for the simulation.

The initially unstable disk model passed through several phases of evolution \citep[see also][]{pickett2003, mejia2005}.  The initial disk is unstable to the growth of axisymmetric structures. During this axisymmetric phase, 
which lasts several ORPs, the disk cools and contracts, the contraction is small radially but dramatic in the vertical direction. Around 3-4 ORPs the instabilities begin grow to non-linear amplitudes. When this condition is met, the disk undergoes a strong burst of GIs that predominantly manifest themselves in one or two discrete global spiral modes.  The disk expands violently during the burst phase, producing a significant rearrangement of the disk's mass distribution on a timescale of a few ORPs. This is then followed by a period of several ORPs where heating temporarily washes out some of the nonaxisymmetry in the disk.  Finally the disk settles into a quasi-steady, long-lived asymptotic state of sustained GI-activity over a large part of the disk, with an overall balance of heating and cooling.

Simulations were run with four different angular resolutions, with grid sizes in $(r,\phi,z)$ of  $(j_{max}, k_{max}, l_{max}) =$ (512,64,64), (512,64,128), (512,64,256), and (512,64,512). Because of the large computational overhead associated with starting each simulation from $t = 0$, the $l_{max} =$ 128 simulation was run through the axisymmetric, burst and transition phases. Simulations using $l_{max} =$ 64, 256, and 512 all begin by interpolating the $l_{max} =$ 128 simulation to higher or lower azimuthal resolution at 9.6 ORPs, near the end of the transition
phase. Although this strategy saves a large amount of computational time, it limits the scope of the analysis to the asymptotic phase and it introduces a perturbation to the disk which takes several ORPs to subside. By 12 ORPs, the perturbation has dissipated and all four disks have transitioned to the asymptotic state.  All simulations were 
run through $\sim 18$ ORPs.

Figure \ref{Final_Q} shows $Q$, as a function of radius, at $t \sim$ 18 ORP, while table \ref{tbl:ams} lists $Q_{avg}$, the value of $Q$ of for the asymptotic disk, time-averaged of 12 -- 19 ORP and spatially averaged over 10 -- 40 AU, as a function of radius for each resolution,

@@

\begin{equation}
A_m = \frac{(a_m^2 + b_m^2)^{1/2}}{\pi\int\rho_o \varpi {\rm d}\varpi {\rm d}z},
\end{equation}
where
\begin{subequations}
\begin{align}
a_m &= \int \rho \cos(m\phi)\varpi {\rm d}\varpi {\rm d}z {\rm d}\phi,\\
b_m &=\int \rho \sin(m\phi)\varpi {\rm d}\varpi {\rm d}z {\rm d}\phi.
\end{align}
\end{subequations}

Here  $\rho_o$ is the axisymmetric (zero-order)  component of the density. In what follows, $A_m$ is averaged over time to suppress fluctuations on the dynamic time scale due to the chaotic nature of GIs in the asymptotic  phase.
In particular, $\langle A_{tot} \rangle$ will represent the power over the radial range 10 -- 40 AU, integrated from $ m = 2$ to $l_{max}/2$ and averaged over the time range 12 -- 19 ORPs. The Fourier amplitudes cannot be accurately measured at resolutions smaller than $l_{max}/2$ \citep{shannon1984}, therefore $A_m$ is limited to modes $m \le l_{max}/2$.  
Because the central star is artificially fixed to the grid center, the $m = 1$ mode cannot be be realistically treated; it is excluded from $\langle A_{tot} \rangle$.

@@

Previous studies have shown that disks in the asymptotic state have power at all resolvable $m$-values  \citep[e.g.][]{mejia2005,boley2006}.  Table \ref{tbl:ams} lists $\langle A_{tot} \rangle$ along with $\langle A_{2-7} \rangle  /  \langle A_{tot} \rangle$,  the fraction of the total power falling into low-order ($m = 2$ -- 7) modes, for each resolution. While the total power is approximately independent of resolution, $\langle A_{tot} \rangle = 1.90 \pm 0.07$, 
the distribution of this power shows marked differences. In particular, the fractional power in lower order modes decreases monotonically with increasing resolution.  Over three quarters of $\langle A_{tot} \rangle$ resides
in modes $m =$ 2 -- 7 for the $l_{max} = 64$ simulation, while approximately half resides in these modes when $l_{max} =$ 512.

The shift of power from lower to higher order modes as the resolution of the computational grid is increased arises from the fact that, while $\langle A_{tot} \rangle$ is approximately independent of $l_{max}$, {\bf SM: Need some explanation as to why this is the case.} the number of degrees of freedom available for this power, $l_{max}/2$, increases linearly with the resolution. Thus power is naturally spread from lower order to higher order modes as the available $m$-values increases. Indeed, even if the disk is dominated by one or two low-order modes at the time of outburst, this power cascades to higher-order modes \citep[for example][]{laughlin1997, laughlin1998,laughlin1996}.

Given the resolution dependence of how power is distributed,  we must examine what resolution is sufficient to properly capture the evolution of these disks. Table \ref{tbl:ams} shows that while the fractional power in lower order modes decreases with each doubling of $l_{max}$, the amount of change in going from $l_{max} = 256$ to 512 
is significantly less than the change going from $l_{max} = 64$ to 128, with a strong suggestion that this diminution will not proceed much further with higher azimuthal resolution. Indeed, a  Richardson extrapolation {\bf (add reference) SM: suggest Press et. al} \citep{press1992} of the tabulated values of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ suggests a limiting value of  $\langle A_{2-7} \rangle / \langle A_{tot} \rangle \sim$ 0.43 for increasingly higher resolutions. This is displayed graphically in figure 4, where the fractional power in lower order modes is plotted against the  normalized grid 
spacing for each value of $l_{max}$. The grid spacing  is normalized such that spacing of the $l_{max}$ = 512 grid
equals unity.  Convergence occurs as the normalized spacing goes to zero. The Richardson extrapolation value  of $\langle A_{2-7}\rangle  / \langle A_{tot}\rangle  = 0.43$ is plotted at a normalized grid spacing  equal to zero. More than half of the modal power in a converged disk resides in higher-order modes, modes that help define a state of gravitoturbulence. {\bf SM: I'm not sure this is accurate, gravitoturbulence isn't really turbulence (at least as it's normally defined) and the high order modes aren't really turbulence either} The fact that the $l_{max} =$ 512 simulation has $\langle A_{2-7} \rangle / \langle A_{tot} \rangle$ near the convergence value lends some confidence in the use of the $l_{max} =$ 512 model for the studie of effective $\alpha$.

@@

Coherent density structures in a disk produce gravitational torques. Low-order modes in $\Delta\rho / \rho$  have longer wavelengths and, if coherent in radius, have longer lever arms to produce torques. On the other hand, high-order modes have relatively short wavelengths which, particularly if they lack coherence, produce more localized effects. The high-order modes tend to produce pockets of local GI activity, which can cancel each other
over the whole disk, {\bf SM: I think there's a sentence missing here explaining that higher resolution leads to more of the amplitude being concentrated in high order modes, while the total amplitude stays the same.} leading to the expectation that  gravitational torques in the disk will decrease as $l_{max}$ is increased. This can be tested by examining the torque contributions from  various Fourier modes.

The torque {\bf C} on a cylindrical surface of the disk at radius $\varpi$ can be obtained by integrating the stress tensor $T$ over the surface of the cylinder \citep{lyndenbell1972}, i.e.,
\begin{equation}
{\rm {\bf C}} = \int {\bf r} \times T \cdot {\rm d}S.
\end{equation}
If the stress tensor includes only gravitational stresses, the surface integral can be replaced with the volume integral
\begin{equation}
{\rm \bf C} = \int \rho {\bf r} \times \nabla \Phi {\rm d}V,
\end{equation}
where $\Phi$ is the gravitational potential.  Here we are interested only in the $z$-component of torque,  
\begin{equation}
\label{torque1}
{\rm \bf C}_Z = \int \rho \frac{\partial\Phi} {\partial \phi} {\rm d}V,
\end{equation}
as only this component drives mass and angular momentum transfer. The torque contribution from each mode $m$ can be calculated by replacing $\rho$ in Equation (5) with the density distribution reconstructed from a single Fourier component
\begin{equation}
\rho_m = a_{\phi m} \cos(m\phi) + b_{\phi m} \sin(m\phi),
\end{equation}
where $a_{\phi m} =\slantfrac{1}{\pi} \int \rho \cos(m\phi){\rm d}\phi$ and $b_{\phi m} =\slantfrac{1}{\pi} \int \rho \sin(m\phi){\rm d}\phi$, and only the gravitational potential produced by the mass distribution given by $\rho_m$ is included in $\Phi$.

Figure 5 {\bf SM: Two things about this figure. Are you sure you removed the m=1? You would have had to have written your own software or significantly changed what I gave you to do this. Also I think the amplitude of the torques are not right if we're really considering a 1$M_\odot$ star. These numbers look like the ones for a 0.5$M_\odot$ star} displays $\langle {\rm \bf C}_{Ztot} \rangle$, the time-averaged gravitational torque summed over all disk modes (excluding $m=1$), along with the torques summed over low-order modes, $\langle {\rm \bf C}_{Z(2-n)} \rangle$, where $n = $ 2, 3, 4, and 5, as a function of radius and azimuthal resolution. 
{\bf (Note to all: the version of figure 5 included with this draft  shows the torques
time-averaged over for four different time-frames.  I did this because I wanted to see if the disk was 
"settling down" with time.  I think the figure suggests  that the 12-14 ORP disk is not
yet in an asymptotic state - or further from it than at later times - I'm not sure this is
a significant difference.  Looking forward to figure 7, it can be argued that the
disk is reasonably settled, in a statistical sense, shortly after 12 ORP.
Dick has told me he only believes the 12-18 ORP result - Dick, I hope 
I'm not misquoting you).  
} 
The scale of the $y$-axis in each subplot is the same, with the exception of the $l_{max} = 64$ simulation. 
In this case the scale is doubled.  Two features are immediately apparent: 1) the total gravitational torque is smaller for higher resolution simulations, and 2) the total torque is dominated by low-order modes for all azimuthal resolutions. Both reflect the fact that a while higher fraction of the total power in larger $l_{max}$ simulations is in higher order modes, these modes make only minimal contributions to the gravitational torque. As resolution increases, non-axisymmetric power is shifted from low-order global modes, which produce relatively large gravitational torques, to high-order local modes, which tend to cancel each other out when integrated over the whole disk.
